Comparing Generative Pre-trained Transformer Models to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6939 Deaths in Sierra Leone from 2019-2022

### 2025-05-18 manuscript/manuscript.tex
Working on child results

Of all 6939 records, GPT-4 (0.61 PCCC) had the highest individual performance followed by GPT-3.5 (0.56 PCCC), InSilicoVA (0.44 PCCC), and InterVA-5 (0.44 PCCC) (Figure \ref{fig-pccc-csmf}). GPT-3.5 and GPT-4 had improvements ranging from 0.14-0.18 PCCC over InSilicoVA and InterVA-5, while GPT-4 slightly improved over GPT-3.5 by 0.05 PCCC. Population level performances were similar for all models (0.74-0.79 CSMF).

Figure \ref{fig-pccc} shows the PCCC performance across age groups. GPT-4 had the best individual performance for adult and neonatal records (0.64 and 0.58 PCCC), while GPT-3.5 had the best performance for child records (0.54 PCCC) with GPT-4 performing slightly worse (0.51 PCCC). InSilicoVA and InterVA-5 performed the worse for adult and child records ($\leq$0.5 PCCC), while GPT-3.5 performed the worse for neonatal records (0.42 PCCC).

Across age ranges, all models followed a similar pattern in individual performance (Figure \ref{fig-pccc-age}. PCCC rose for records with ages from 0 to 27 days, dropped drastically for 1-5 months, rose again from 6 months to 14 years, and slowly declined for ages 15 to 69 years. The highest and lowest performances were observed for ages 10-29 years and 1-11 months respectively. Performances varied more across models for ages 0 days to 5 years, while less variation was seen between GPT-3.5 and GPT-4, as well as InSilicoVA and InterVA-5, from 5 to 69 years.

@@

Figure \ref{fig-pccc-cod-adult} shows model performance by PCCC across 17 adult CODs excluding suicide, due to a low 3 (\textless 1\%) records. GPT-4 had the highest individual performance for 10 of 17 CODs (0.36 to 0.99 PCCC), InSilicoVA for 2 CODs (0.72 and 0.84 PCCC), and GPT-3.5 for 5 CODs (0.44-0.94 PCCC). The lowest PCCC values were InterVA-5 between 0 to 0.79 for 8 of 17 CODs (InSilicoVA and GPT-3.5 within 0.05 for cancers and unspecified infections respectively), InSilicoVA between 0.01 to 0.41 for 6 of 17 CODs (InterVA-5 within 0.05 for other cardiovascular diseases and diarrhoeal diseases), and GPT-3.5 at 0.39 for tuberculosis and 0.54 for road and transport injuries. Chronic respiratory diseases had the largest differences between minimum and maximum PCCC at 0.93. Maternal conditions, and road and transport injuries, had higher PCCC values above or at 0.67 (with the exception of GPT-3.5 for road and transportation injuries as an outlier), where there were small differences between minimum and maximum PCCC below or at 0.2. Ischemic heart disease and unspecified infections had PCCC values below or at 0.65, where there were small differences between minimum and maximum PCCC at or below 0.14. Each adult COD had between 27 (\textless 1\%) to 799 records (21\%) with malaria (n=799, 21\%), other noncommunicable diseases (n=414, 11\%), and unspecified infections (n=389, 10\%) making 42\% of all records, and other CODs between 1-9\% of all records.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{fig-pccc-cod-adult.pdf}
\caption{Model performance for adult records by COD.}\label{fig-pccc-cod-adult}
\end{figure}

@@

For the 8 child CODs (excluding congenital anomalies due to small sample size of 1 (\textless 1\%) record), the highest PCCC observed was for GPT-4 for 4 of 8 CODs, GPT-3.5 for 3 of 8 CODs (InSilicoVA within 0.05 for other infections), and InSilicoVA for pneumonia (Figure \ref{fig-pccc-cod-child}). The lowest PCCC across the 8 child CODs was observed for InSilicoVA for epilepsy, leukaemia, and other noncommunicable diseases and malaria, InterVA-5 for pneumonia, nutritional deficiencies, and other infections. GPT-3.5, InSilicoVA, and InterVA-5 all had similar PCCC values greater than 0.8 as the lowest PCCC for injuries. Epilepsy, leukaemia, and other noncommunicable disease had the largest difference between minimum and maximum PCCC at 0.68, while Ill-defined had the next largest difference at 0.67 with InSilicoVA and InterVA-5 at 0 PCCC. Malaria and other infections had low PCCC values below 0.58 across models with small differences between minimum and maximum PCCC at 0.18 and 0.14 respectively. Injuries and diarrhoeal diseases had high PCC values above 0.6 across models with small differences between minimum and maximum PCC at 0.14 and 0.3 respectively. Each of the 8 child CODs had between 11 (\textless 1\%) to 1382 (52\%) records with malaria (n=1382, 52\%), other infections (n=667, 25\%), and diarrhoeal diseases (n=175, 7\%) making 84\% of all records. Nutritional deficiencies had less than 1\% of all records, while other CODs had between 1-3\% of all records.

>>>>

Of all 6939 records, GPT-4 (0.61 PCCC) had the highest individual performance followed by GPT-3.5 (0.56 PCCC), InSilicoVA (0.44 PCCC), and InterVA-5 (0.44 PCCC) (Figure \ref{fig-pccc-csmf}). GPT-3.5 and GPT-4 had improvements ranging from 0.14-0.18 PCCC over InSilicoVA and InterVA-5, while GPT-4 slightly improved over GPT-3.5 by 0.05 PCCC. Population level performances were similar for all models (0.74-0.79 CSMF). Figure \ref{fig-pccc} shows the PCCC performance across three age groups (adult, child, and neonate). GPT-4 had the best individual performance for adult and neonatal records (0.64 and 0.58 PCCC), while GPT-3.5 had the best performance for child records (0.54 PCCC) with GPT-4 performing slightly worse (0.51 PCCC). InSilicoVA and InterVA-5 performed the worse for adult and child records ($\leq$0.5 PCCC), while GPT-3.5 performed the worse for neonatal records (0.42 PCCC). Across age ranges, all models followed a similar pattern in individual performance (Figure \ref{fig-pccc-age}). PCCC rose for records with ages from 0 to 27 days, dropped drastically for 1-5 months, rose again from 6 months to 14 years, and slowly declined for ages 15 to 69 years. The highest and lowest performances were observed for ages 10-29 years and 1-11 months respectively. Performances varied more across models for ages 0 days to 5 years, while less variation was seen between GPT-3.5 and GPT-4, as well as InSilicoVA and InterVA-5, from 5 to 69 years.

@@

Figure \ref{fig-pccc-cod-adult} shows model performance by PCCC across 17 adult CODs excluding suicide, due to a low 3 (\textless 1\%) records. GPT-4 had the highest individual performance for 10 of 17 CODs (0.35 to 0.99 PCCC), GPT-3.5 for 5 CODs (0.43-0.94 PCCC), and InSilicoVA for 2 CODs (0.71 and 0.84 PCCC). InterVA-5 had the lowest performance for 8 of 17 CODs (0-0.79 PCCC), InSilicoVA for 6 CODs (0.01-0.41 PCCC), and GPT-3.5 for 2 CODs (0.38 and 0.53 PCCC). GPT-3.5/4 models improved over InSilicoVA/InterVA-5 the most for chronic respiratory diseases (0.74-0.94 PCCC difference), and the least for Malaria (0.09-0.17 PCCC difference). All models performed well for maternal conditions (0.79-0.99 PCCC), while unspecified infections, malaria, and ill-defined CODs had low performing models (\textless 0.5 PCCC). GPT-4 had large performance improvements compared to all other models for cancers (+0.25-0.36 PCCC), stroke (+0.27-0.45 PCCC), and diarrhoeal diseases (+0.37-0.51 PCCC), while GPT-3.5 had similarly large improvements for liver and alcohol related diseases (+0.27-0.52 PCCC).

\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{fig-pccc-cod-adult.pdf}
\caption{Model performance for adult records by COD.}\label{fig-pccc-cod-adult}
\end{figure}

@@

Figure \ref{fig-pccc-cod-child} presents individual performances for each of the models by 8 child CODs (excluding congenital anomalies due to having only 1 (\textless 1\%) record). the highest PCCC observed was for GPT-4 for 4 of 8 CODs, GPT-3.5 for 3 of 8 CODs (InSilicoVA within 0.05 for other infections), and InSilicoVA for pneumonia (). The lowest PCCC across the 8 child CODs was observed for InSilicoVA for epilepsy, leukaemia, and other noncommunicable diseases and malaria, InterVA-5 for pneumonia, nutritional deficiencies, and other infections. GPT-3.5, InSilicoVA, and InterVA-5 all had similar PCCC values greater than 0.8 as the lowest PCCC for injuries. Epilepsy, leukaemia, and other noncommunicable disease had the largest difference between minimum and maximum PCCC at 0.68, while Ill-defined had the next largest difference at 0.67 with InSilicoVA and InterVA-5 at 0 PCCC. Malaria and other infections had low PCCC values below 0.58 across models with small differences between minimum and maximum PCCC at 0.18 and 0.14 respectively. Injuries and diarrhoeal diseases had high PCC values above 0.6 across models with small differences between minimum and maximum PCC at 0.14 and 0.3 respectively. Each of the 8 child CODs had between 11 (\textless 1\%) to 1382 (52\%) records with malaria (n=1382, 52\%), other infections (n=667, 25\%), and diarrhoeal diseases (n=175, 7\%) making 84\% of all records. Nutritional deficiencies had less than 1\% of all records, while other CODs had between 1-3\% of all records.
